\documentclass[11pt]{article}

% ------------------------------------------------------------------
% Packages
% ------------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{hyperref}
\usepackage[numbers]{natbib}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage[margin=1in]{geometry}

% ------------------------------------------------------------------
% Theorem environments
% ------------------------------------------------------------------
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

% ------------------------------------------------------------------
% Convenience macros
% ------------------------------------------------------------------
\newcommand{\St}{S_t}
\newcommand{\Pt}{P_t}
\newcommand{\Mt}{M_t}
\newcommand{\Bt}{B_t}
\newcommand{\Gt}{G_t}
\newcommand{\Vt}{V_t}
\newcommand{\Et}{E_t}
\newcommand{\calT}{\mathcal{T}}
\newcommand{\calP}{\mathcal{P}}
\newcommand{\taint}{\tau}
\newcommand{\prov}{\pi}
\newcommand{\SYS}{\textsf{SYS}}
\newcommand{\USER}{\textsf{USER}}
\newcommand{\TOOL}{\textsf{TOOL}}
\newcommand{\TOOLauth}{\textsf{TOOL}_{\textsf{auth}}}
\newcommand{\TOOLunauth}{\textsf{TOOL}_{\textsf{unauth}}}
\newcommand{\WEB}{\textsf{WEB}}
\newcommand{\SKILL}{\textsf{SKILL}}

% ------------------------------------------------------------------
\title{Noninterference Theorem for Indirect Prompt Injection\\in Agentic AI}
\author{}
\date{}

\begin{document}
\maketitle

% ==================================================================
\begin{abstract}
Agentic AI systems routinely ingest untrusted content---emails, web pages,
documents, and tool outputs---in the same overall workflow that carries system
prompts and user intent. Indirect prompt injection exploits this mixed-control
environment by embedding instructions inside otherwise task-relevant data,
allowing adversarial content to influence tool calls, plan updates, memory
writes, or final answers. We formalise the desired security property as a
\emph{noninterference} condition adapted from information-flow
control~\cite{goguen1982}: adversarial changes in untrusted content should not
change control-relevant behaviour.

The paper makes two contributions that must be interpreted separately. First,
we present a formal theorem for a system-level enforcement architecture based on
a typed intermediate representation, taint tracking, and guard checks before
control-relevant actions. Second, we report an executable empirical prototype
implemented in this repository: a contaminated-input guarded runtime, a concrete
structured declassification mechanism, and an end-to-end benchmark comparing an
unguarded baseline, a guarded system without declassification, and a guarded
system with structured declassification. On a saved benchmark of seven attacked
scenario families and seven clean counterparts, run with three seeds, the
unguarded baseline exhibits 1.000 attack success rate and 1.000 unsafe action
rate, while the guarded system with structured declassification achieves 0.000
attack success, 0.000 unsafe action rate, and 0.857 attacked task completion,
matching its clean task completion on this benchmark. These results support a
narrow but concrete empirical claim: explicit declassification can recover
usefulness that is otherwise lost under strict guarding, without reintroducing
unsafe actions in the implemented scenarios. They do not establish a general
production-security guarantee.
\end{abstract}

% ==================================================================
\section{Introduction}\label{sec:intro}

Agentic AI systems increasingly combine language-model reasoning with external
tools, persistent memory, browsing, code execution, and long-horizon task
state. Because these systems consume untrusted content such as retrieved web
pages, user-uploaded documents, and unauthenticated tool outputs alongside
trusted instructions, they are vulnerable to \emph{indirect prompt injection}:
an attacker hides new instructions inside data channels that the agent was
supposed to merely read~\cite{greshake2023,crowdstrike2025,esecplanet2025}.
The attack is dangerous precisely because it blurs the boundary between
\emph{content} and \emph{control}. A malicious string in an email body can
redirect a tool call; a poisoned retrieval result can alter a plan; a tainted
memory write can corrupt later decisions.

We argue that the right formal lens for this problem is
\textbf{noninterference}. In the language of information-flow
control~\cite{goguen1982}, the relevant question is whether adversarial
variation in untrusted inputs can influence low-level, control-relevant outputs.
In an agent setting, those outputs include tool invocations, plan updates,
persistent memory writes, and final user-visible answers. If a system permits
untrusted content to change those outputs arbitrarily, then data has become
control.

This repository began as a formal-security prototype. Its original strength was
conceptual clarity: it identified noninterference as the right specification
and proposed taint-aware filtering as the enforcement mechanism. Its empirical
story, however, was weaker than the formal story. In particular, earlier
``guarded'' evaluations depended too heavily on shortcut-style reasoning about
clean versus contaminated inputs, rather than demonstrating that the guarded
system could operate on the same contaminated state as the baseline.

This paper presents the repository in its updated form. The theorem remains a
statement about a \emph{system-level architecture}, not about the internal
reasoning of a language model. The empirical contribution is now narrower but
more credible: the repository contains an executable contaminated-input runtime,
an explicit declassification mechanism, and a saved end-to-end benchmark that
measures both security and utility.

\paragraph{Formal claim versus empirical claim.}
The central formal claim is conditional: if tainted content is excluded from the
control-relevant dependency set under the stated invariants, then control
outputs are noninterfering with respect to adversarial changes in untrusted
input. The central empirical claim is correspondingly narrower: in the
implemented benchmark, a guarded runtime with structured declassification blocks
the benchmarked attacks while recovering task completion lost by a strict
non-declassifying guard.

\paragraph{Contributions.}
\begin{enumerate}[label=(\roman*)]
  \item A formal model of agent state with a typed intermediate representation,
        taint tracking, and a trust lattice that distinguishes trusted,
        unauthenticated, and verified information sources.
  \item A noninterference theorem for system-level action selection, stated and
        proved under explicit invariants (Sections~\ref{sec:model} and
        \ref{sec:theorem}).
  \item An implemented contaminated-input guarded runtime that enforces policy
        at tool-call, plan-update, memory-write, and final-answer control
        points.
  \item A concrete structured declassification mechanism that validates typed
        fields and promotes only accepted values into verified state.
  \item A saved end-to-end benchmark, with raw artifacts and aggregate tables,
        comparing unguarded, guarded-without-declassification, and
        guarded-with-declassification variants on both security and utility.
\end{enumerate}

% ==================================================================
\section{Background: Indirect Prompt Injection as a Control-Plane Threat}
\label{sec:background}

Indirect prompt injection has been documented as a practical threat to
LLM-integrated systems because the model often processes user intent and
third-party content in the same context~\cite{greshake2023,crowdstrike2025}.
CrowdStrike describes the risk as environmental poisoning: the adversary need
not interact directly with the system if they can control content that the
system later reads~\cite{crowdstrike2025}. eSecurity Planet similarly notes
that many systems lack a hard separation between explicit instructions and
external content, which lets hostile data reshape tool use, workflow control,
and data access decisions~\cite{esecplanet2025}.

Benchmarks such as InjecAgent~\cite{zhan2024injecagent} and
BIPIA~\cite{yi2023bipia} were important because they established that indirect
prompt injection is measurable rather than merely anecdotal. They also made
clear that prompt-only defenses remain probabilistic. SpotLight, for example,
can sharply reduce attack success in some settings but still leaves the model
with access to the untrusted content and therefore does not establish a hard
system-level guarantee~\cite{hines2024spotlight}.

Our approach differs in emphasis. Rather than relying solely on prompt design,
we treat indirect prompt injection as an \emph{architectural information-flow
problem}. The goal is not to convince the model to ignore the attack, but to
ensure that untrusted content cannot directly determine control-relevant output
unless it first passes through an explicit declassification mechanism.

% ==================================================================
\section{Model Definitions}\label{sec:model}

\input{definitions}

% ==================================================================
\section{Noninterference Theorem}\label{sec:theorem}

\input{proof}

% ==================================================================
\section{Interpreting the Theorem as a System Specification}
\label{sec:discussion}

\subsection{What the theorem does and does not say}

A skeptical reader may restate the theorem as follows: if one removes tainted
information from the control-relevant dependency set before action selection,
then the resulting action-selection function is invariant to changes in that
removed information. We agree with this characterization. The theorem is not
valuable because it is surprising; it is valuable because it specifies exactly
\emph{which architectural condition} is sufficient for the desired security
property, and exactly \emph{which assumptions} must hold for that condition to
be meaningful.

This is the right level of abstraction for an agent-security theorem. The proof
is not about the internals of a neural network. It is about a system boundary:
which information channels are permitted to influence control decisions, and
which are not. In that sense, the theorem functions less like a claim about
emergent model behaviour and more like a specification for a taint-aware
control architecture.

\subsection{The utility problem and the need for declassification}

A strict reading of noninterference creates an immediate practical difficulty.
Many useful tasks require an agent to read untrusted content and then act on
something extracted from it. If all action-relevant influence from untrusted
content is forbidden, then the system may be secure but useless. This is the
core utility problem.

The repository now addresses that problem explicitly through declassification.
The implemented strategy reduces untrusted text to bounded typed fields,
validates them against policy and schema checks, and promotes only accepted
values into verified state. Unsafe spans are quarantined rather than discarded,
so the system can retain them for inspection without allowing them to steer
control decisions. The theorem itself does not prove that a given
\emph{declassification policy} is correct; it only makes clear where such a
policy must sit in the architecture.

\subsection{Threat-model boundaries}

Neither the theorem nor the implementation should be read as a universal
security result. The current scope excludes malicious user intent, compromised
system prompts, and arbitrary mislabeling of authenticated channels. It also
does not prove that every future declassification strategy will be sound.
Instead, it provides a framework within which such strategies can be made
explicit, tested, and compared.

% ==================================================================
\section{Implemented Empirical Prototype}
\label{sec:implementation}

The empirical layer in this repository now operates on contaminated inputs
rather than shortcutting through pre-cleaned guarded runs. The prototype is
model-agnostic and stateful. Trusted inputs, untrusted content, and tool
outputs are ingested through explicit channels. Provenance and trust labels are
stored as state nodes. Every proposed control-relevant action is then evaluated
by the guard before it may affect the plan, memory, action execution, or final
answer.

\paragraph{Guarded runtime.}
The guarded runtime enforces policy at four principal control points: tool calls
or action execution, plan updates, memory writes, and final-answer emission.
Tainted evidence can still exist in the runtime and can still be transformed
into candidate facts, but it cannot directly justify control-relevant actions.
For example, tainted tool or action proposals are blocked; tainted plan updates
are rejected; tainted final answers are repaired to a safe template; and
tainted memory writes are quarantined rather than promoted to verified memory.
The runtime also logs each allowed or blocked decision together with its
evidence, blocking rules, and repair behaviour.

\paragraph{Structured declassification.}
The repository now implements a concrete declassification strategy based on
structured schema extraction. Untrusted text is parsed into candidate fields,
validated against typed field specifications and instruction-detection rules,
and only then promoted into verified facts. The implementation currently
supports typed extraction for values such as email addresses, dates, prices,
URLs, and product identifiers. It rejects instruction-like payloads and retains
unsafe spans as quarantined context rather than permitting them to influence
later actions.

\paragraph{Why this matters empirically.}
This architecture closes the main circularity that affected the earlier
prototype. The guarded system is no longer ``guarded'' merely because it is
handed clean inputs. It is guarded because it consumes the same contaminated
state as the baseline and enforces explicit policy within that state.

% ==================================================================
\section{End-to-End Evaluation}
\label{sec:eval}

\subsection{Design}

The repository's main empirical evidence is now an end-to-end local benchmark
implemented under \texttt{noninterference\_eval}. The benchmark compares three
system variants:
\begin{enumerate}[label=(\arabic*)]
  \item \textbf{baseline}: an unguarded runtime that applies proposed actions
        without taint-aware restriction;
  \item \textbf{guarded\_no\_declass}: a guarded runtime that blocks or
        repairs tainted control actions but does not recover trusted facts via
        declassification;
  \item \textbf{guarded\_structured\_declass}: the guarded runtime plus the
        structured declassification strategy.
\end{enumerate}

Each scenario is executed in both attacked and clean form for each variant.
The canonical saved benchmark run in this repository uses three seeds
(0, 1, 2), seven attacked scenarios, and seven clean counterparts. The seven
attacked scenario families are:
\begin{enumerate}[label=(\alph*)]
  \item markdown embedded instruction override in email summarization;
  \item hostile HTML retrieval poisoning with paraphrased terminal-style
        instructions;
  \item two-step persistence poisoning through memory;
  \item JSON tool-output poisoning;
  \item a field-level declassification-bypass attempt;
  \item obfuscated override instructions embedded in a markdown code block;
  \item authenticated-tool paraphrase poisoning.
\end{enumerate}

\subsection{Metrics}

The updated benchmark reports both security and usefulness. Security metrics
include attack success rate, unsafe action rate, security failure rate, blocked
malicious action rate, persistence-poisoning success, and indirect prompt
influence. Utility metrics include attacked and clean task completion,
attacked and clean task-token match, benign-action preservation, overblocking,
and runtime overhead versus baseline. Raw attacked traces, clean traces,
per-scenario tables, per-seed summaries, and aggregate mean--standard
deviation tables are saved under the run directory.

\subsection{Results}

Table~\ref{tab:main-results} reports the aggregate results from the saved run.
Most core outcome metrics are deterministic across the three seeds in this
benchmark; runtime overhead varies modestly because the runtime is lightweight
and the absolute timings are small.

\begin{table}[t]
\centering
\caption{Saved end-to-end benchmark results from
\texttt{noninterference\_eval/results/e2e/benchmark\_20260327T025110Z}.}
\label{tab:main-results}
\begin{tabular}{lcccc}
\toprule
\textbf{Variant} & \textbf{Attack Success} & \textbf{Unsafe Action} & \textbf{Security Failure} & \textbf{Attacked Task Completion} \\
\midrule
baseline & $1.000 \pm 0.000$ & $1.000 \pm 0.000$ & $1.000 \pm 0.000$ & $0.286 \pm 0.000$ \\
guarded\_no\_declass & $0.000 \pm 0.000$ & $0.000 \pm 0.000$ & $0.429 \pm 0.000$ & $0.286 \pm 0.000$ \\
guarded\_structured\_declass & $0.000 \pm 0.000$ & $0.000 \pm 0.000$ & $0.000 \pm 0.000$ & $0.857 \pm 0.000$ \\
\bottomrule
\end{tabular}
\end{table}

A fuller picture appears when task preservation is included. The unguarded
baseline has clean task completion of $0.857 \pm 0.000$, but attacked task
completion drops to $0.286 \pm 0.000$ because successful injections redirect
behaviour. The guarded system without declassification removes malicious tool
execution and unsafe actions, but it also collapses clean task completion to
$0.286 \pm 0.000$, demonstrating the expected utility loss of a strict guard.
The structured-declassification variant restores clean and attacked task
completion to the same level, $0.857 \pm 0.000$, while maintaining zero attack
success and zero unsafe actions on this benchmark.

\begin{table}[t]
\centering
\caption{Utility and operational trade-offs in the saved benchmark.}
\label{tab:utility-results}
\begin{tabular}{lcccc}
\toprule
\textbf{Variant} & \textbf{Clean Task Completion} & \textbf{Benign-Action Preservation} & \textbf{Indirect Prompt Influence} & \textbf{Runtime Overhead vs Baseline} \\
\midrule
baseline & $0.857 \pm 0.000$ & $0.143 \pm 0.000$ & $1.000 \pm 0.000$ & $0.000 \pm 0.000\%$ \\
guarded\_no\_declass & $0.286 \pm 0.000$ & $0.857 \pm 0.000$ & $1.000 \pm 0.000$ & $-5.137 \pm 22.008\%$ \\
guarded\_structured\_declass & $0.857 \pm 0.000$ & $1.000 \pm 0.000$ & $0.000 \pm 0.000$ & $79.946 \pm 45.207\%$ \\
\bottomrule
\end{tabular}
\end{table}

These numbers support three conclusions. First, the baseline is genuinely
vulnerable in the implemented scenarios; this is not a tautological benchmark.
Second, simple guarding without declassification is insufficient as a practical
solution because it preserves safety at the cost of capability. Third,
structured declassification materially changes that trade-off in the current
benchmark by recovering legitimate task behaviour while preserving safe control
outcomes.

\subsection{What the results do not prove}

The saved benchmark is still modest in scale. Seven scenario families are not a
substitute for internet-scale agent traffic, and local benchmark timings are
not a substitute for deployment measurements. The result should therefore be
read as evidence that the architecture is executable and that the utility gap
can be narrowed in a concrete implemented setting. It should not be read as a
claim that indirect prompt injection is solved for arbitrary tools, arbitrary
agent frameworks, or arbitrary authenticated channels.

% ==================================================================
\section{Limitations and Future Work}
\label{sec:limitations}

The updated repository is stronger empirically than the initial prototype, but
its evidence remains bounded.

\begin{enumerate}[label=(\arabic*)]
  \item \textbf{Small benchmark.} The current end-to-end evaluation covers
        seven scenario families rather than a large heterogeneous corpus.
  \item \textbf{Local runtime.} The implementation is a reusable research
        runtime, not a direct integration into a production agent framework.
  \item \textbf{Authenticated-channel assumptions.} The theorem and runtime
        both remain sensitive to how tool outputs are labeled and promoted.
        Mislabeling a hostile channel as trusted would undermine the guarantee.
  \item \textbf{Overhead interpretation.} Runtime overhead is measured within a
        lightweight local harness and should not be interpreted as deployment
        latency.
  \item \textbf{Future declassification work.} Only one concrete
        declassification strategy is implemented today. Broader strategy
        families should be benchmarked, stress-tested, and formally analyzed.
\end{enumerate}

Future work should therefore focus on larger and more diverse scenario
collections, framework-level integrations, stronger authenticated-channel
assumptions, richer declassification strategies, and adversarial testing aimed
specifically at the verification pathway.

% ==================================================================
\section{Conclusion}\label{sec:conclusion}

Indirect prompt injection is fundamentally a boundary-failure problem: data from
untrusted channels is allowed to behave like control. The noninterference
perspective is useful because it states the desired security condition in a
precise and architecture-facing way. Under explicit invariants, tainted content
must not be permitted to influence control-relevant outputs.

The updated repository now supports that formal perspective with a more honest
empirical artifact. It implements a contaminated-input guarded runtime, a
concrete structured declassification mechanism, and a saved end-to-end
benchmark that compares baseline, guarded-without-declassification, and
guarded-with-declassification systems on both security and utility. In that
benchmark, strict guarding alone is safe but utility-poor, whereas structured
declassification recovers the benchmark's clean task completion while
preserving zero unsafe actions.

That result is encouraging, but it is deliberately limited. The repository now
provides a credible starting point for empirical work on agentic noninterference;
it does not justify a blanket claim that indirect prompt injection has been
solved in general.

% ==================================================================
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
